---
title: "7. Prespecified scoring and the frozen one-shot audit"
author: "Jibeom Ko"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. Prespecified scoring and the frozen one-shot audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
suppressPackageStartupMessages(library(pdactrace))
set.seed(42)
```

# Goal

The vignettes so far audit evidence that is already in hand. This one
covers the other half of the package: ranking tissue-to-blood candidates
from stage-resolved effect tables, and then **committing to a decision
rule before the destination cohort is opened**.

The motivation is a negative result. Two PDAC serum audits were
prespecified, hashed, externally timestamped, and opened exactly once.
Both terminated without producing a performance statistic, because the
frozen candidate panel was not quantifiable at the required depth in the
public serum deposits. The record is in `audit/PROVENANCE.md` in the
source repository. That is why the runner below has a terminal
`INCONCLUSIVE` state that is deliberately *not* a failure of the marker.

Everything in this vignette runs on simulated data, so no download is
needed and the numbers are reproducible from the seeds shown.

# Step 1 -- A simulated multi-cohort, multi-omic benchmark

`sb_simulate()` generates a stage-resolved effect table, a blood
detection table, and the feature-level truth used only for evaluation.
The `translation` scenario plants coherent tissue-only decoys and
inflammation-associated blood decoys, so a method that merely rewards
strong tissue signal can be caught out.

```{r simulate}
sim <- sb_simulate(n_features = 400L, scenario = "translation", seed = 1L)
vapply(sim, nrow, integer(1))
```

`sb_validate_effects()` is the input contract. It normalises the five
required columns, drops non-finite estimates, and refuses duplicated
feature/cohort/modality/contrast rows rather than silently averaging
them.

```{r validate}
effects <- sb_validate_effects(sim$effects)
str(effects[, c("feature", "cohort", "modality", "contrast", "estimate")])
unique(effects$modality)
```

# Step 2 -- Depth-adjusted blood observability

A detection in a shallow cohort is more informative than the same
detection in a cohort where nearly everything is seen, so raw detection
frequency is not used directly.

```{r observability}
head(sb_observability(sim$detection), 3)
```

# Step 3 -- Rank candidates

`stagebridge()` combines eight components into a weighted geometric
score. The geometric form matters: a candidate with one near-zero
component cannot be rescued by the other seven, which is the intended
behaviour for a translation claim.

`target_relation` must be declared **before** looking at the
destination data. `"concordant"` asks for tissue and blood to move the
same way; `"inverse"` asks for the opposite; `"agnostic"` scores
magnitude only.

```{r rank}
ranked <- stagebridge(
  sim$effects,
  detection = sim$detection,
  target_relation = "concordant"
)
head(ranked[, c(
  "feature", "stagebridge_score", "score_rank",
  "pareto_front", "pareto_tier"
)], 5)
```

Two tiers are reported and they are deliberately separate.
`pareto_front` is exact non-dominance over the eight components;
`score_percentile_tier` is a score-only cut. A candidate can lead on
score while being dominated, and that disagreement is information.

```{r pareto}
components <- ranked[, c(
  "target", "anchor", "evidence", "replication",
  "concordance", "observability", "coverage", "source"
)]
table(sb_pareto_fronts(components))[1:4]
```

# Step 4 -- Does the ranking survive dropping a cohort?

```{r loco}
loco <- sb_loco_stability(sim$effects, sim$detection, top_k = 20L)
head(loco, 3)
```

`worst_loco_rank` is the number to read. A candidate whose median rank
is 2 but whose worst leave-one-cohort-out rank is 150 is carried by a
single cohort.

# Step 5 -- Comparators, scored honestly

The package ships the comparators rather than only its own score, so
that claims of improvement can be checked instead of asserted.

```{r comparators}
scores <- sb_method_scores(sim$effects, sim$detection)
truth <- sim$truth$truth[match(scores$feature, sim$truth$feature)]
methods <- setdiff(names(scores), "feature")

evaluation <- data.frame(
  method = methods,
  auroc = vapply(methods, function(m) {
    sb_binary_auc(truth, scores[[m]])
  }, numeric(1)),
  auprc = vapply(methods, function(m) {
    sb_average_precision(truth, scores[[m]])
  }, numeric(1)),
  recall_top20 = vapply(methods, function(m) {
    sb_topk_recall(truth, scores[[m]], k = 20L)
  }, numeric(1)),
  row.names = NULL
)
evaluation[order(-evaluation$auroc), ]
```

On this scenario StageBridge has the **lowest AUROC of the eight
methods** while ranking second on AUPRC and on top-20 recall. That is
the expected shape of the trade-off, not a bug: the score is tuned to
put translatable candidates at the very top of a shortlist, and it pays
for that with global ranking quality. `observed_rank_product` beats it
on all three summaries here. Read this as a demonstration of the
evaluation harness, not as a claim that one method dominates -- a single
simulated scenario cannot settle that, and the scenario argument exists
precisely so that assumptions can be violated on purpose.

`sb_multiblock_plsda_scores()` is the matched-sample comparator. Unlike
everything above it needs the same observations, in the same order, in
every block, so it does not apply to unpaired cohort-level effect
tables. It is shown here for completeness.

```{r mbplsda, eval = requireNamespace("mixOmics", quietly = TRUE)}
outcome <- factor(rep(c("Normal", "Tumor"), each = 12))
dn <- list(paste0("S", seq_len(24)), paste0("F", seq_len(8)))
rna <- matrix(stats::rnorm(24 * 8), nrow = 24, dimnames = dn)
protein <- rna + stats::rnorm(24 * 8, sd = 0.5)
head(sb_multiblock_plsda_scores(list(RNA = rna, Protein = protein), outcome), 3)
```

# Step 6 -- The frozen one-shot audit

Everything above is exploratory and re-runnable. `sb_one_shot()` is not.
It takes a frozen panel, a frozen direction per feature, and a rule
fixed in advance, then returns exactly one of `PASS`, `FAIL` or
`INCONCLUSIVE`.

```{r rule}
str(sb_default_one_shot_rule())
```

The gate stack runs first: group ontology and counts, broad candidate
coverage, top-10 and top-20 panel coverage, and complete-panel
quantifiability. The statistical endpoint runs **only** if every gate
passes.

```{r fixture}
make_cohort <- function(delta, n_covered = 320L, seed = 20260719L) {
  frozen <- data.frame(
    feature = sprintf("G%03d", 1:320),
    frozen_direction = rep(c(1L, -1L), length.out = 320L)
  )
  panel <- data.frame(
    feature = frozen$feature[1:20],
    frozen_direction = frozen$frozen_direction[1:20],
    rank_weight = 1 / log2(2:21)
  )
  samples <- c(
    sprintf("CP_%02d", 1:20), sprintf("CN_%02d", 1:20),
    sprintf("HC_%02d", 1:20)
  )
  grid <- expand.grid(
    gene = frozen$feature[seq_len(n_covered)], sample = samples,
    KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
  )
  grid$group <- sub("_.*$", "", grid$sample)
  grid$log2_abundance <- withr::with_seed(
    seed, stats::rnorm(nrow(grid), 8, 1)
  )
  if (delta != 0) {
    hit <- grid$group == "CP" & grid$gene %in% panel$feature[1:10]
    direction <- panel$frozen_direction[match(grid$gene, panel$feature)]
    grid$log2_abundance[hit] <-
      grid$log2_abundance[hit] + delta * direction[hit]
  }
  list(
    abundance = grid, frozen = frozen, panel = panel,
    groups = list(
      case = "CP", primary_reference = "CN", secondary_reference = "HC"
    ),
    counts = c(CP = 20L, CN = 20L, HC = 20L)
  )
}

# Fewer resamples than the locked rule, to keep the vignette quick.
quick_rule <- function() {
  rule <- sb_default_one_shot_rule()
  rule$permutation_replicates <- 200L
  rule$bootstrap_replicates <- 200L
  rule
}

audit <- function(...) {
  cohort <- make_cohort(...)
  sb_one_shot(
    cohort$abundance, cohort$frozen, cohort$panel,
    cohort$groups, cohort$counts, quick_rule()
  )
}
```

A cohort with a real planted signal reaches `PASS`:

```{r pass}
passed <- audit(delta = 1.2)
c(status = passed$status, auc = round(passed$primary$auc, 3))
```

The same cohort with the signal removed reaches `FAIL`. The endpoint
ran and did not meet the criteria -- this is a statement about the
biology:

```{r fail}
failed <- audit(delta = 0)
c(status = failed$status, runner_entered = failed$statistical_runner_entered)
```

Now the case that both real PDAC audits hit. The signal is still
present, but only 50 of the 320 frozen candidates are quantifiable, so
the coverage gate halts execution before the endpoint. The result is
`INCONCLUSIVE`: the test never ran, and nothing has been learned about
the marker:

```{r inconclusive}
halted <- audit(delta = 1.2, n_covered = 50L)
c(status = halted$status, runner_entered = halted$statistical_runner_entered)
halted$failed_gates
```

Collapsing that third outcome into `FAIL` would have reported a
biological conclusion that the data cannot support. Keeping it separate
is the point of the design.

# Why a calibration check matters

A gate stack strict enough never to pass anything would produce the same
`INCONCLUSIVE` results on real data while proving nothing. The package's
test suite therefore drives the runner to all three terminal states and
checks that AUROC rises monotonically with the size of the planted
signal (`tests/testthat/test-sb-one-shot.R`). `PASS` is reachable, so
the real-data terminations are attributable to assay coverage rather
than to an over-conservative rule.

# Scope

The two recorded PDAC audits, their accessions and their terminal states
are documented in `audit/PROVENANCE.md`. The lock manifests hash paths
relative to the root of the archived working tree and are covered by
external RFC 3161 timestamps, so they verify only inside that archive;
the copies under `audit/locks/` are a disclosure record. Regenerating a
manifest would destroy the property it exists to establish.

```{r session_info}
sessionInfo()
```
